The tracing of trade

### 2021-01-29 Manuscript/Manuscript.tex
manuscript corrected

\documentclass[review]{elsarticle}
\usepackage{hyperref,lineno}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\modulolinenumbers[5]

\newcommand{\memo}[2]{\textcolor{#1}{#2}}
\newcommand{\maria}[1]{\memo{red}{MC: #1\\}}
\newcommand{\xavi}[1]{\memo{magenta}{XRC: #1\\}}

\journal{Journal of Archaeological Method and Theory}

\bibliographystyle{model2-names.bst}\biboptions{authoryear}

@@

\begin{frontmatter}

%\title{The markings of the trade: exploring the patterns of olive oil production in Roman Baetica}
\title{The tracing of trade: exploring the patterns of olive oil production and distribution from Roman Baetica}

\author[ceipacadress]{María Coto-Sarmiento\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{mcotsar@gmail.com}

\author[edadress,ubadress]{Xavier Rubio-Campillo}

\address[ceipacadress]{Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, Universitat de Barcelona, Barcelona, Spain}
\address[edadress]{School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK}
\address[ubadress]{DIDPATRI, Universitat de Barcelona, Passeig de la Vall d'Hebrón, 171, Barcelona, Spain}

\begin{keyword}
Roman Empire; amphora production; Dressel 20; dissimilarity index; Roman provinces
\end{keyword}

@@

%Our case of study has been focused on the production processes located in \textit{Baetica} province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. \textit{Baetica} became an important production and distribution centre during the Roman Empire. However, it remains under debate about how this province was organised and whether it could be possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance. The analysis explores how the quantitative approach provides a useful tool for the interpretation of the economic processes. Finally, results pretend to highlight the organisation of Baetican olive oil production in the Roman Empire linked to the differences observed in the archaeological evidence.

\end{abstract}

\end{frontmatter}

\section{Introduction}

@@

%Material culture is one of the most frequent indicators of trade in the archaeological record. In archaeology, they allow us to highlight a part of the mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016} 
%spread of goods and ideas had a enormous impact during the Roman Empire. The progressive exploitation of communication networks allowed a major interaction between communities
%This can be seen by the fact that material culture is found in different regions along the Mediterranean as a frequent indicator of trade exchange in the archaeological record.

@@

Advances in the research of the Roman studies have currently led an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.
%\xavi{more diverse que qué?} 
The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of Roman production
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

This work focuses on the debate about the dynamics of the distribution of olive oil market in Roman provinces. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing stamps amphorae finding in the provinces.

The finding of amphoric stamps in different provinces has allowed us to know the importance of the distribution of olive oil within the Empire. Despite having archaeological evidence, no consensus has been reached: How was the economical relation between production and consumption centres? Did they follow an economical pattern to distribute olive oil in those provinces?

This paper aims to study the olive oil market connection between provinces by calculating the similarity of stamps. Specifically, our work pretends to detect regional processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to understand the pattern of olive oil production linked to amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/general_map}
\caption{Overview of the sites analysed in this paper. The colour defines the three different regions under study (\textit{Baetica} as producer; \textit{Germania} and \textit{Britannia} as consumers) while the size of each dot highlights the number of Dressel 20 amphoric stamps found on each of these sites}

@@

In particular, we study the distribution of amphoric stamps to identify a correlation between geographical distance and similarity. Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of stamps, b) stamps located in close workshops share similar traits and c) low mobility of amphoric stamps to other regions: stamps always stay in the same region.

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}.  A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here, we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

@@

\section{The Amphoric production of Baetica}

The landscape of the \textit{Baetican} province was one of the best regions to face the increased demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were built near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/baetica}
\caption{Distribution of stamps found in Dressel 20 workshops along the Guadalquivir riverbank. The size of each point depends on the number of stamps found on each site}

@@

%Some stamps show a more specific chronology while the majority of them display a large activity of production that it can be difficult for specifying an accurate chronology in some cases. This could be due to two reasons; firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible; secondly, Dressel 20 was produced during almost three centuries with apparently few changes \citep{berni_piero_chapter_2017}.

\section{A potential indicator of Roman economy: Dressel 20 stamps}

@@

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{figs/dressel20}
\caption{Dressel 20 were mostly marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure}

@@

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion.

Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to initial forms of a name but they can also appear complete.

@@

%Additionally, a large percentage of Dressel 20 containers were not stamped thus complicating the interpretation

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}. 
%In any case, the use of these stamps is a good proxy to explore the system of olive oil production and trade developed in Baetica.

@@

%\subsubsection{Jaccard distance}
%The dataset was analysed using a statistic method as Jaccard distance. This method allows to measure the dissimilarity by calculating the presence of sets (CITAR). In our case, Jaccard distance was used to compute the mutual presence of traits in the amphora stamps but it does not consider the number of absences. A comparison was done with the distance of the workshops to identify whether there was an association between stamps and spatial distance amongst workshops.

\section{Consumption centres: Britannia and Germania}

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes.

This led to a gradual change in the economic and social structures the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}

@@

%The importance of this olive oil trade is revealed by the massive amount of Dressel 20 amphorae found across the Roman provinces. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.\xavi{me parece repetitivo, porque unos párrafos antes ya hablabas de lo mismo. Quizás unificar texto aquí o allá?} LO HE PUESTO MÁS ARRIBA!

@@

An explanation for the interest to export olive oil for the military garrison in the Roman provinces is still unclear. The origin of this exportation may be related to two aspects: a) cultural consumption whose olive oil is used by cultural reasons such as group identification or by habit and b) economical and political reasons to create a redistributive system mechanism to supply military camps \citep[69-70]{carreras_britannia_1998}.

@@

\subsection{Britannia}

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced on the region as the landscape and climate of the British Isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}.

@@

%At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}\xavi{repetitivo? fusionar párrafos}.

@@

%The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns\xavi{estas 2 frases repiten cosas ya explicadas}.

@@

There are no written records explaining how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep[45]{funari_economic_2005}.

@@

\subsection{Germania}

@@

%This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}

@@

Previous studies suggested that the supply to the German limes was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia})~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

\section{Material and Methods}

@@

To study the amphoric stamps, we used the data collected in the CEIPAC database of amphoric epigraphy from different places~\citep{remesal_centro_2015} (see the database here \url{http://romanopendata.eu}). The CEIPAC dataset contains over 50.000 epigraphy records found on different types of amphorae, mostly from \textit{Monte Testaccio}.

%\xavi{podrías mover esta parte a los 2 párrafos anteriores y explicar la bbdd brevemente antes de empezar a hablar de la Bética (porque es info que también aplica a los otros 2 casos de estudio no?}.

%This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. 
%\xavi{estos 2 párrafos explican objetivos, que deberían ir en la intro y no aquí, donde deberías empezar a hablar directamente de la BBDD}

\subsection{Production centres: Baetica province}

The database allowed us to retrieve 3798 stamps found in Dressel 20 amphorae recovered from workshops within \textit{Baetica} province. Each of these stamps ha detailed records on the site where it was found, the inscribed code as well as its spatial coordinates. The stamps with incomplete information were discarded, thus finishing with a sample of 987 stamps from 81 sites and displaying 130 different code.

%Approximately 70 \% of stamps cannot be tested due to the fragmentation of the dataset. Consequently, any stamps with incomplete information were discarded and not integrated into our dataset.

@@

The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter \textit{Hispalis}), \textit{Cordubensis} (currently C\'ordoba, hereafter \textit{Corduba}) and \textit{Astigi} (currently Écija, Sevilla, hereafter \textit{Astigi}) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

It is important to mention that some workshops exhibited the same geographical coordinates while they were catalogued differently as workshops. For instance, two nearby amphora workshops such as Tesorillo de Doña Mencia and Doña Mencia were catalogued as a different name in spite of having the same geographical coordinates. Our work considered this particularity in the archaeological dataset but some biases can occur in the analysis.

@@

%\xavi{no lo veo; recuerda que para el rollo bayesiano tenías una línea recta al hacedr el log de los 2 ejes, no solo 1, y este plot no está construído así no?}.
%\maria{son dos ejes, no? ¿aquí no se ve línea recta? Lo mismo si lo hago de puntitos...}

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequencystamp.pdf}
\caption{Histogram on a log scale with base 10. X axis is represented by the number of stamps and Y axis is the frequency of workshop. The distribution is widely diverse with most workshops having only one stamp}
\label{stamps}
\end{figure}

The distribution of code stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 code stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this sample size (267 and 146 code stamps). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exceptions of 2 large workshops in \textit{Hispalis} (La Catria and Arva). On these two workshops a comparatively large amount of code stamps were found (29 different code stamps on each of them).

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequency}
\caption{Distribution of the number of different code stamps (X axis) for each \textit{conventus} (Y axis). Each dot corresponds to a workshop sorted by different areas. Colours are represented by areas divided into \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{frequency}
\end{figure}

\subsection{Consumption centres: Britannia and Germania}

@@

The output was a dataset of 4271 stamps found in sites belonging to \textit{Britannia} (total: 2219 stamps, analysed: 1765 stamps ) and \textit{Germania} (total: 2052 stamps, analysed: 1621 stamps). Some stamps from \textit{Britannia} and \textit{Germania} were discarded for the analysis for incomplete information.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/britannia}
\caption{Sites in \textit{Britannia} where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. A majority of stamps have been found in garrisons related to Hadrian's and Antonine walls}
\label{britannia}
\end{figure}

%We analysed a dataset of 2052 stamps placed in \textit{Germania}. All the data was also compiled by CEIPAC database. As previously before, the centres were selected with more or equal than five stamps.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/germania}
\caption{Sites in Germania where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. Most sites are located along the German \textit{limes}}
\label{germania}
\end{figure}

\subsection{Measuring the dissimilarity}

%quizás habría que hablar también del filtrado de códigos que se hizo en python porque antes había 3783 stamps pero se filtró por el número de letras...
%\xavi{lo del filtrado lo puedes poner de supplementary information con referencia al antiquity}

The approach proposed here aimed at exploring links between production and consumption centres through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn index, we first calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop.

If two workshops had similar stamp code distribution then the distance index would be close to 0 whereas sites with completely unrelated stamp codes would give a distance close to 1.

\subsection{Hierarchical clustering}

The Morisita-Horn index was used to generate a dissimilarity matrix containing the distance value between each pairwise site based on their code stamps. Hierarchical clustering was applied to this matrix in order to group sites with similar stamps codes distributions. The algorithm was selected to cluster similar groups in order to analyse the relationship between groups of sites and the distribution of similar stamp codes. The results were visualised using a dendrogram to detect groups of sites sharing similar stamp codes.

\section{Results}

\subsection{Production centres: Baetica Province}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%

\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{lscape}
\usepackage{rotating}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system

\usepackage{natbib}
%\bibliographystyle{plainnat}
\bibliographystyle{agsm}

%\bibliographystyle{model2-names.bst}%\biboptions{authoryear}
%\bibliographystyle{spbasic}
%\bibliographystyle{spphys}
%\bibliographystyle{spmpsci}
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
\journalname{Archaeological and Anthropological Sciences}

@@

\title{The tracing of trade%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
\subtitle{Exploring the patterns of olive oil production and distribution from Roman Baetica}

%\titlerunning{Short form of title}        % if too long for running head

\author{Mar\'ia Coto-Sarmiento       \and Xavier Rubio-Campillo}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Mar\'ia Coto-Sarmiento \at Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, Universitat de Barcelona, Barcelona, Spain
\email{mcotsar@gmail.com}            
\emph{mcotsar@gmail.com} 
%of F. Author  %  if needed
           \and
\\
Xavier Rubio-Campillo \at
School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK \at
DIDPATRI, Universitat de Barcelona, Passeig de la Vall d'Hebr\'on, 171, Barcelona, Spain             
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

@@

%Insert your abstract here. Include keywords, PACS and mathematical subject classification numbers as needed.
\keywords{Roman Empire \and amphora production \and Dressel 20 \and dissimilarity index \and Roman provinces}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}

@@

@@

Advances in Roman studies have currently led to an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.

The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of Roman production
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

This work focuses on the debate about the dynamics of the distribution of olive oil market in Roman provinces. Amphoric production in \textit{Baetica} province will help us identify links between production centres and consumption centres in order to detect an economic pattern in the commerce by analysing amphoric stamps finds in the provinces.

The finding of amphoric stamps in different provinces has allowed us to know the importance of the distribution of olive oil within the Empire. Despite having archaeological evidence, no consensus has been reached: How was the economic relation between production and consumption centres? Did they follow an economic pattern to distribute olive oil in those provinces?

This paper aims to study the olive oil market connection between provinces by calculating the similarity of amphoric stamps. Specifically, our work aims to detect regional processes focused on a commercial product (i.e. olive oil) from a specific province \citep{isaksen_network_2006}.

%We want to understand the pattern of olive oil production linked to amphora workshops and stamps used to mark them. 
%We focus here on exploring the economic relation between stamps and amphora production and distribution centres.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{general_map}
\caption{Overview of the sites analysed in this paper. The colour defines the three different regions under study (\textit{Baetica} as producer; \textit{Germania} and \textit{Britannia} as consumers) while the size of each dot highlights the number of Dressel 20 amphoric stamps found on each of these sites}

@@

%In particular, we study the distribution of amphoric stamps to identify a correlation between geographical distance and similarity.

Based on this assumption, we proposed three hypotheses: a) we can identify a correlation between spatial distance and the distribution of amphoric stamps, b) stamps located in close workshops share similar traits and c) low mobility of amphoric stamps to other regions: stamps always stay in the same region.

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}. Quantitative approach provides one way to measure the similarity between amphoric stamps. Here, we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance.

%Stamps will be used to identify economic patterns by analysing their similarity.

For example, if workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

@@

\section{The Amphoric production of Baetica}
\label{sec:1}

The landscape of the \textit{Baetican} province was one of the best regions to face the increased demand for olive oil across the Roman provinces. For this reason, the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third century AD~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were built near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).The proximity distance between workshops was relatively close which allowed for more efficient distribution.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{baetica}
\caption{Distribution of stamps found in Dressel 20 workshops along the Guadalquivir riverbank. The size of each point depends on the number of stamps found on each site}

@@

\section{A potential indicator of Roman economy: Dressel 20 stamps}
\label{sec:2}

@@

\begin{figure}[htp]
	\centering
\includegraphics[scale=0.5]{dressel20}
\caption{Dressel 20 were mostly marked with stamps of three letters called \textit{tria nomina}}
\label{amphora}
\end{figure}

@@

The use of amphoric stamps can be a good proxy to explore the system of olive oil production and trade developed in \textit{Baetica} province. These stamps are typically found on handles but they were also imprinted on rims and the amphora body~\citep{millet_anforas_1998}. The content information of the stamps had not a unique criterion. They showed a code with different letters and forms. Stamps often displayed a code of three letters and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}. The three letters correspond to initial forms of a name but they can also appear complete.

@@

The stamp codes are interpreted based on three main ideas: content (olive oil), context (amphora workshop) and subject (individuals involved in the production). On the one hand, it seems that stamps could have been identified as the landowner of the olive groves \citep{rodriguez_economioleicola_1977}. On the other hand, they could also belong to the owner of the workshops where the amphorae were made or even to a group of amphora workers~\citep{berni_millet_epigrafianforica_2008}.

@@

\section{Consumption centres: Britannia and Germania}
\label{sec:3}

The creation of new provinces required the Roman Empire the arrival of resources through Mediterranean and Atlantic routes. This effectively led to a gradual change in the economic and social structures when the trade networks that supplied the Roman army became more complex.

Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}

@@

@@

An explanation for the interest to export olive oil for the military garrison in the Roman provinces is still unclear. The origin of this exportation may be related to two aspects: a) consumption based on cultural reasons (i.e. social identification or habit  by a group of consumers) and b) economical and political reasons to create a redistributive system mechanism to supply military camps \citep[69-70]{carreras_britannia_1998}.

@@

\subsection{Britannia}
\label{sec:4}

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. The local population did not consume this product and it was not produced on the region as the landscape and climate of the British Isles was not suitable for the growth of olive trees~\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}.

@@

@@

@@

There are no written records explaining how this redistribution of essential goods was organised in \textit{Britannia}, but the archaeological evidence suggests that cities may have been the central nodes of this local trade network~\citep{funari_economic_2005,orengo_seeds_2016}.

@@

\subsection{Germania}
\label{sec:4}

@@

@@

Previous studies suggested that the supply to the German \textit{limes} was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for \textit{Germania} and \textit{Britannia})~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

\section{Material and Methods}
\label{sec:5}

@@

To study the amphoric stamps, we used the data collected in the CEIPAC database of amphoric epigraphy from different places~\citep{remesal_centro_2015} (see the database here \url{http://romanopendata.eu}). The CEIPAC dataset contains over 50.000 epigraphy records found on different types of amphorae, mostly from \textit{Monte Testaccio}. Most stamps used in the analysis did not have a specific chronology or were not available in the dataset. This fact did that we had to discard the possibility to analyse a distribution of stamps per chronologies.

In order to enable reproducibility of the results, the selected datasets and source codes are available on the following GitHub repository: \url{https://github.com/Mcotsar/Baetica\_stamps}.

\subsection{Production centres: Baetica province}
\label{sec:5}

The database allowed us to retrieve 3798 stamps found in Dressel 20 amphorae recovered from workshops within \textit{Baetica} province. Each of these stamps has detailed records at the site where it was found, the inscribed code as well as its spatial coordinates. The stamps with incomplete information were discarded, thus finishing with a sample of 987 stamps from 81 sites and displaying 130 different codes. Query dataset from Baetica province can be found on the following GitHub repository: \url{https://github.com/Mcotsar/Baetica_stamps/tree/main/Databases/Baetica}.

@@

The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter \textit{Hispalis}), \textit{Cordubensis} (currently C\'ordoba, hereafter \textit{Corduba}) and \textit{Astigi} (currently \'Ecija, Sevilla, hereafter \textit{Astigi}) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

It is important to mention that some workshops exhibited the same geographical coordinates while they were catalogued differently as workshops. For instance, two nearby amphora workshops such as Tesorillo de Do\~na Mencia and Do\~na Mencia were catalogued as a different name in spite of having the same geographical coordinates. Our work considered this particularity in the archaeological dataset but some biases can occur in the analysis.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{frequencystamp.pdf}
\caption{Histogram on a log scale with base 10. X axis is represented by the number of stamps and Y axis is the frequency of workshop. The distribution is widely diverse with most workshops having only one stamp}
\label{stamps}
\end{figure}

The distribution of codes in stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this sample size (267 and 146 stamps respectively). The workshops of the three \textit{conventus} show a similar pattern on the stamp frequency distribution with the exceptions of 2 large workshops in \textit{Hispalis} (La Catria and Arva). On these two workshops a comparatively large amount of stamp codes were found (29 different codes on each of them).

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{frequency}
\caption{Distribution of the number of different codes in stamps (X axis) for each \textit{conventus} (Y axis). Each dot corresponds to a workshop sorted by different areas. Colours are represented by areas divided into \textit{Hispalis} (red), \textit{Astigi} (green) and \textit{Corduba} (blue)}
\label{frequency}
\end{figure}

\subsection{Consumption centres: Britannia and Germania}
\label{sec:5}

@@

The output was a dataset of 4271 stamps found in sites belonging to \textit{Britannia} (total: 2219 stamps, analysed: 1765 stamps ) and \textit{Germania} (total: 2052 stamps, analysed: 1621 stamps). The stamps with incomplete information were also not taken into account for the analysis.

Query datasets from both provinces can be found on the following GitHub repository: Britannia~ \url{https://github.com/Mcotsar/Baetica_stamps/tree/main/Databases/Britannia} and Germania~ \url{https://github.com/Mcotsar/Baetica_stamps/tree/main/Databases/Germania}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{britannia}
\caption{Sites in \textit{Britannia} where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. A majority of stamps have been found in garrisons related to Hadrian's and Antonine walls}
\label{britannia}
\end{figure}

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{germania}
\caption{Sites in Germania where Dressel 20 amphoric stamps have been found. Each dot is a site while its size shows the sample size of stamps found on the site. Most sites are located along the German \textit{limes}}
\label{germania}
\end{figure}

\subsection{Measuring the dissimilarity}
\label{sec:5}

The approach proposed here aimed at exploring links between production and consumption centres through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of stamps. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}. In our case, stamps from different sites will be used as a proxy to measure their dissimilarity by analysing their code.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn index, we first calculated the number of times that each stamp code appeared in an amphora workshop. 
This method allowed us to bear in mind the number of frequency of stamps for each workshop. For example, if two workshops had similar stamp code distribution then the distance index would be close to 0 whereas sites with completely unrelated stamp codes would give a distance close to 1.

\subsection{Hierarchical clustering}
\label{sec:5}

The Morisita-Horn index was used to generate a dissimilarity matrix containing the distance value between each pairwise site based on their codes stamped in amphorae. Hierarchical clustering was applied to this matrix in order to group sites with similar stamp codes distributions. The algorithm was selected to cluster similar groups in order to analyse the relationship between groups of sites and the distribution of similar stamp codes. The results were visualised using a dendrogram to detect groups of sites sharing similar stamp codes.

\section{Results}
\label{sec:6}
